---
title: "Introduction to SomaPlotr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to SomaPlotr}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/SomaPlotr-",
  fig.align = "center",
  dpi = 80
)
```


This vignette is intended for analysts that are interested in learning how to 
visualize SomaScan data with the `SomaPlotr` R package. `SomaPlotr` contains
numerous plotting functions that are specifically designed to identify and display 
patterns in SomaScan data. These functions provide a fast and simple mechanism 
for producing high-quality graphics without extensive programming or data 
visualization experience. 

This vignette will walk through examples of each plotting function in the package.
`SomaPlotr` is built around [ggplot2](https://ggplot2.tidyverse.org),
so if you are familiar with `ggplot2`, you can extend this package by further
modifying or customizing the plots as desired. 
 

-------------


## Setup

`SomaPlotr` can be loaded with a simple call to `library()`:

```{r load-package}
library(SomaPlotr)
```

In addition to `SomaPlotr`, this vignette will require the following packages
for example data sets, data wrangling functions, and plotting utilities.
The plots in this vignette will be generated using the example ADAT provided in
[SomaDataIO](https://somalogic.github.io/SomaDataIO/):

```{r other-packages, message = FALSE}
library(dplyr)
library(ggplot2)
library(SomaDataIO)
data <- SomaDataIO::example_data
```


----------------

## CDF and PDF Plots

Cumulative distribution function (CDF) and probability density function (PDF) 
plots are frequently used visualization methods when analyzing RFU values 
obtained from SomaScan. These plots can be generated in various forms using the 
functions provided in `SomaPlotr`. 

### PDF Plots

A PDF plot is created from a numeric (double) vector of, usually, RFU values
with `plotPDF()`:

```{r pdf, fig.width = 6, fig.height = 4}
plotPDF(data$seq.8468.19)
```

This plot has a fairly long tail, which highlights a common feature of SomaScan
data in RFU space. As a consequence, we typically advise to plot (and analyse)
RFU data in `log10()` space:

```{r pdf-log10, fig.width = 6, fig.height = 4}
plotPDF(log10(data$seq.8468.19))
```

The above plot displays a characteristic bimodal distribution, suggesting that
some underlying structure may be present. `plotPDFbyGroup()` creates a PDF
plot split by a grouping variable of (usually) metadata, e.g. `Sex`.
This function differs from `plotPDF()` in that it requires a data frame as input. 

For the plot below, the variable `Sex` will be used to stratify the groups:

```{r pdf-group, fig.width = 6, fig.height = 5}
# Clean up data by removing missing 'Sex' values
# and log10() transform
df_sex <- filter(data, !is.na(Sex)) |> log10()

# Generate PDF plot for analyte of interest
plotPDFbyGroup(df_sex, apt = "seq.8468.19", group.var = Sex)
```

Using `plotPDFlist()`, a PDF plot can be generated from any arbitrary _named_
list of numeric vectors, with one smoothed kernel density curve per list element:

```{r pdf-list, fig.width = 6, fig.height = 5}
list_seq <- list(
  dil_20 = data$NormScale_20,
  null   = withr::with_seed(101, rnorm(nrow(data), mean = 1, sd = 0.2))
)

plotPDFlist(
  list_seq,
  fill = TRUE,
  main = "20% Scale Factors vs 1.0 Null Distribution",
  x.lab = "Scale Factor"
)
```


### CDF Plots

Similarly, CDF plots can be created with a suite of functions that serve as 
counterparts to the PDF plotting functions displayed above: `plotCDF()`,
`plotCDFbyGroup()`, and `plotCDFlist()`. These functions are implemented like the
PDF plotting functions, and use the same type of inputs:

```{r cdf-plots, eval = FALSE}
plotCDF(data$seq.8468.19)

plotCDFbyGroup(df_sex, apt = "seq.8468.19", group.var = Sex)

plotCDFlist(list_seq)
```


--------------


## Concordance Plots  
  
Plots illustrating the concordance between two continuous variables (e.g. RFU 
values from SomaScan analytes) can be generated with `plotConcord()`.

These plots accept two numeric vectors as input; one for `x`, and one for `y`:

```{r concord-vec, fig.width = 6, fig.height = 4}
x <- df_sex$seq.3045.72
y <- withr::with_seed(1, x + rnorm(length(x), sd = 0.1))  # add gaussian noise
plotConcord(x, y)
```

However, `plotConcord()` will also accept a 2-column data frame, like so:

```{r concord-df, eval = FALSE}
df_2col <- data.frame(x = x, y = y)
plotConcord(df_2col)
```

------------------

## Volcano Plots

Volcano plots comparing p-value vs. fold change can be generated using 
`plotVolcano()`.

`plotVolcano()` requires a 2-column data frame of fold change values and 
p-values as input. These data types are not available in our example ADAT, so 
they will each be simulated below:

```{r volc-df}
withr::with_seed(101, {
  fc1 <- sort(runif(500, -2.5, 0))   # Z-scores as fc
  fc2 <- sort(runif(500, 0, 2.5))    # Z-scores as fc
  p1  <- pnorm(fc1)
  p2  <- pnorm(fc2, lower.tail = FALSE)
  p   <- jitter(c(p1, p2), amount = 0.1)
  p[p < 0] <- runif(sum(p < 0), 1e-05, 1e-02)   # floor p < 0 after jitter
  fc_df <- data.frame(fc = c(fc1, fc2), p = p)
})
```

This data frame can now be used as input into `plotVolcano()`:

```{r volc-plot, fig.width = 6, fig.height = 5}
# lower p-value cutoff than default
plotVolcano(fc_df, fc, p, cutoff = 0.1)
```

The `labels=` and `identify=` arguments can be used to label the points:

```{r volc-html-df, fig.width = 6, fig.height = 5}
target_map <- getTargetNames(getAnalyteInfo(data))
target_map

# add random SeqId rownames to `fc_df`
fc_df <- set_rn(fc_df, withr::with_seed(1, sample(getAnalytes(data), nrow(fc_df))))
# map rownames to target labels
fc_df$target_names <- unlist(target_map, use.names = TRUE)[rownames(fc_df)]

plotVolcano(
  fc_df,
  fc,
  p,
  cutoff   = 0.005,
  labels   = target_names,   # add target labels to points
  identify = TRUE
)
```

With large significance levels (i.e. higher cutoff), labels can quickly
become cluttered, overlapping, and difficult to distinguish.
An interactive HTML-based volcano plot can be created to explore the fold
change and p-values of each individual point using `plotVolcanoHTML()`.
This function can use the same input data as `plotVolcano()`, but uses 
[plotly](https://plotly.com/r/)
under the hood to generate a hovering menu that can be used to
interactively investigate each point:

```{r volc-html-plot, fig.width = 6, fig.height = 6, fig.align = "center"}
plotVolcanoHTML(
  fc_df,
  fc,
  p,
  cutoff        = 0.05,
  labels = target_names
)
```

-------------------
  
  
## Boxplots

The `boxplotGrouped()` function can be used to plot a response variable (`y`) 
split by the specified grouping variable(s); up to 2 may be used.

Below, boxplots are generated for a single grouping variable, `Sex`.

```{r box-grp-1var, fig.width = 6, fig.height = 5}
boxplotGrouped(
  .data     = df_sex,         # log10-transformed soma_adat
  y         = "seq.8468.19",  # PSA target
  group.var = "Sex",          # grouping variable
  beeswarm  = TRUE            # add beeswarm points
)
```

One additional grouping variable can be added, if desired. For example,
`PlateId` is used as the second variable to split the data by both `Plate`
and `Sex` to investigate possible plate bias by gender:

```{r box-grp-2var, fig.width = 6, fig.height = 5}
boxplotGrouped(
  .data     = df_sex,
  y         = "seq.8468.19",        # PSA target
  group.var = c("Sex", "PlateId")   # 2 splitting variables
)
```

Boxplots with "beeswarm"-style points can be created with `boxplotBeeswarm()`. 
Note that the boxes in the plot below correspond to columns of the data 
frame, _not_ groups within a categorical metadata variable
(as seen in `boxplotGrouped()`).

```{r box-beeswarm, fig.width = 6, fig.height = 5}
boxplotBeeswarm(
  data.frame(seq.10056.5 = data$seq.10056.5,
             seq.10021.1 = data$seq.10021.1) |> log10()
)
```

Notice the small group of low-signalling values that can be seen by adding
"beeswarm" points. These represent buffer control samples and will be
investigated further below.

Lastly, `boxplotSubarray()` can be used to visualize the distribution of all 
analytes, stratified by subarray, each as its own boxplot.
In the SomaScan context, a _subarray_ refers to an individual sample or
row of data.

```{r box-subarray, fig.width = 8, fig.height = 5}
samples <- withr::with_seed(123, sample(rownames(data), 20L))
df_subarray <- data[samples, ]
boxplotSubarray(df_subarray, color.by = "SampleType")
```

In the figure above, each boxplot represents a single sample/subarray/row
(the x-axis is labeled by `SampleId`), and boxes are
colored by sample type (via `SampleType`). 
RFU values (`log10`-transformed by default) for all available analytes are
plotted for each sample.

In addition to the `color.by=` argument, the `apts=` argument can be passed
to highlight specific analytes within each subarray.

```{r box-subarray-apts, fig.width = 8, fig.height = 5}
seqs <- c("seq.8468.19", "seq.3045.72")
boxplotSubarray(df_subarray, color.by = "SampleType", apts = seqs)
```


----------------


## Histograms

`SomaPlotr` provides one histogram plotting function, `plotDoubleHist()`, that
allows the distribution of two numeric vectors to be overlaid for easy 
visual comparison:

```{r double-hist, fig.width = 6, fig.height = 4}
withr::with_seed(123,
  data.frame(
    seq.1234.56 = rnorm(1000, 2, 0.3),
    seq.9876.54 = rnorm(1000, 3, 0.3)
  )
) |> plotDoubleHist()
```


---------------


## Longitudinal Data

Change in subjects across time can be tracked using `plotLongitudinal()`. As the 
name suggests, this function is designed to track RFU measurements in sample 
groups over time. `plotLongitudinal()` requires input data of a different format
than previously described plots. Instead, a measurement must be present for each 
sample type and subject at each time point. To satisfy this requirement, 
additional data will need to be simulated and added to the example data set
we have been using.
This will better emulate the type of longitudinal study data that 
`plotLongitudinal()` is designed to visualize.

```{r long-plot-df}
df_long <- withr::with_seed(123, {
  samples <- sample(df_sex$SampleId, 6L) # Select a subset of samples for the fake study
  data.frame(
    SampleId    = rep(samples, each = 3L),
    TimePoint   = rep(c("0 baseline", "12 mo", "24 mo"), 6L), # Add timepoint measurements
    TissueType  = rep(c("Whole blood", "Plasma", "White blood cells"), each = 3L), # Specify tissue collection
    seq.10021.1 = sample(df_sex$seq.10021.1, 18L)) # Sample RFU measurements for analyte of interest
})
head(df_long, 10L)
```

The longitudinal plot can now be generated:

```{r long-plot, fig.width = 8, fig.height = 4}
plotLongitudinal(
  data  =  df_long,
  y     = seq.10021.1,
  time  = TimePoint,
  id    = SampleId,
  color = TissueType,
  summary.line = NULL   # suppress summary lines
)
```

Lines stemming from each point at `baseline` signify the
change in analyte `seq.10021.1` over time, based on the tissue type
of the collected sample.


--------

## Color Palettes

For examples of all plotting themes and palettes available in this
package, see the themes vignette (`TODO`).


