---
title: "Applying SomaLogic Themes and Palettes"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{Applying SomaLogic Themes and Palettes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/Themes-",
  fig.align = "center",
  dpi = 80,
  fig.width = 7,
  fig.height = 6
)
```


This vignette provides an overview of the color palettes, 
[ggplot2](https://ggplot2.tidyverse.org/) themes, and color scales available in
`SomaPlotr`. These features can be used to generate colorful, polished, and 
visually consistent graphics.

The following packages will be required to run the code in this vignette 
natively:

```{r load-packages}
library(SomaPlotr)
library(ggplot2)
library(gridExtra)
library(withr)
library(scales)
```


# Hexadecimal Color Lists
___

`SomaPlotr` provides lists that define hexadecimal codes for colors used in
SomaLogic branding and marketing materials. A hexadecimal (hex) color is 
specified with the format `#RRGGBB`, where the red (`RR`), green (`GG`), and
blue (`BB`) components of the color are hex values between `00` and `FF`. 
The hexadecimal color lists in `SomaPlotr` are designed to clearly distinguish 
groups and can be applied to a variety of graphics types, with options for 
both color and greyscale graphics.


## Branding Colors

`soma_colors` is a list of hex values corresponding to `r length(soma_colors)` 
colors from the SomaLogic Operating Co. branding color scheme (prior to 2020), in darker 
and more subdued shades than what is used today (in 2023).

```{r soma-colors}
# Display the hex color codes in the list
soma_colors

# Show a preview of all list colors
scales::show_col(unlist(soma_colors))
```

`soma_colors2` contains the exact colors (N = `r length(soma_colors2)`) used 
in _current_ official SomaLogic branding materials, but contains fewer values 
than `soma_colors`:

```{r, fig.width = 4, fig.height = 4}
soma_colors2
scales::show_col(unlist(soma_colors2))
```

Values from these color lists can be directly supplied to visualization 
utilities (like [ggplot2](https://ggplot2.tidyverse.org/)) to designate 
colors for specific graphical elements:

```{r soma-colors-ggplot-example}
# Provide color values for individual plot elements
ggplot(iris, aes(x = Species, y = Sepal.Width)) +
  geom_boxplot(fill = soma_colors2$teal,
               outlier.color = soma_colors2$yellow) +
  geom_jitter(color = soma_colors2$pink)
```

When coerced to a vector, the entire palette can be supplied as an argument to 
base R plotting utilities:

```{r soma-colors-base-R-example, fig.width = 7, fig.height = 5}
hist(iris$Sepal.Length, col = unlist(soma_colors))
```

## Greys and Risk

`SomaPlotr` provides two additional lists of convenient color schemes for 
non-color or risk-stratified graphics. The first, `soma_colors_greys`, 
contains only greyscale values, and is best used for materials (e.g. printed 
documents) where color display is not guaranteed:

```{r soma-colors-greys}
soma_colors_greys
scales::show_col(unlist(soma_colors_greys))
```

The second, `soma_colors_risk`, contains colors commonly used for risk 
designation:

```{r soma-colors-risk}
soma_colors_risk
scales::show_col(unlist(soma_colors_risk))
```


# Color Palettes Generator
___

The color lists described in the previous section contain a finite number of 
values. For graphics involving a large number of groups, these lists may not 
provide enough values for each group. The `palette_soma()` function remedies 
this by by recycling values from the `soma_colors2` list, allowing the user 
to generate a hex color vector of N length.

```{r palette-soma-example}
# This list only contains 8 values
length(soma_colors2)

# This function can create a vector of any length
palette_soma(10)
```

See below for visual displays of `palette_soma()` in action:

```{r palette-soma-grid}
scales::show_col(palette_soma(5))

scales::show_col(palette_soma(10))
```

For example, in the simulated dataset below, there are 15 groups, but 
`soma_colors2` only contains 8 values:

```{r example-dataset}
groups <- paste("Grp", seq(1, 15, 1L))
withr::with_seed(123, {
  n <- 200
  df <- data.frame(seq.1234.56 = rnorm(n, 2, 0.5),
                   Group = sample(groups, n, replace = TRUE)
                   )
})
```

We can use `palette_soma()` to recycle the `soma_colors2` values and generate
a hex vector of length `r length(groups)`. We can then use that vector to 
assign colors (using `ggplot2::scale_fill_manual()`) in a graphic containing 
boxplots:

```{r palette-soma-boxplots, fig.width = 9, fig.height = 5}
pal <- palette_soma(length(groups))

ggplot(df, aes(x = Group, y = seq.1234.56, fill = Group)) +
  geom_boxplot() +
  scale_fill_manual(values = pal) +
  theme(legend.position = "none")
```


# ggplot2 Themes & Scales
___

## Theme

`ggplot2` themes provide a method for easy customization of all non-data 
components of plots (e.g. titles, labels, fonts, background, grid lines, 
legends, and more). `SomaPlotr` uses themes to give plots a consistent look,
as well as save time and effort when finalizing and polishing multiple plots. 
The `theme_soma()` theme applies a predetermined set of modifications to 
each plot for convenience and consistency. These include modifying the x- and 
y-axis labels, removing the (default) grey-filled background, moving the 
figure legend, and more:

```{r default-theme, fig.width=11, fig.height=5}
# Default ggplot2 theme
default_plot <- gg$point
themeSoma_plot <- gg$point + theme_soma()

grid.arrange(default_plot, themeSoma_plot, ncol = 2)
```


## Scales

`SomaPlotr` contains `ggplot2`-compatible scales, both continuous and 
discrete, that can be applied to previously generated plots. See below for 
examples.

```{r scale-color-soma}
gg$point + scale_color_soma()
```

```{r scale-fill-soma}
gg$bar + scale_fill_soma()
```

```{r scale-fill-soma-risk}
gg$box + scale_fill_soma_risk()
```

Other scales include:

  - `scale_continuous_color_soma()`
  - `scale_continuous_color_soma_risk()`
  - `scale_continuous_fill_soma()`
  - `scale_continuous_fill_soma_risk()`
  - `scale_fill_soma_gender()`
